---
title: "Top countries"
subtitle: "Sirius Project - Working Paper 001"
author: "Claude Grasland"
date: "2026-02-24"
date-format: iso
lang: en
format:
  html:
    embed-resources: true
    theme: yeti
    smooth-scroll: true
    fontsize: 0.9em
    toc: true
    toc-depth: 2
    toc-title: "."
    css: custom.css
    crossrefs-hover: false
execute:
  warning: false
  message: false 
  cache: false   
  echo: false
  comment: false
knitr:
  opts_chunk:
    out.width: "100%"
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(knitr)
library(FactoMineR)
library(sf, quietly = T)
library(DT)
#library(mapsf, quietly = T)
#library(visNetwork, quietly=T)
```

## Research question

Which countries are the most mentionned in the news ? What are the variations according to media or time period ?How to aggregate the data from different sources.




## Data preparation

### Load media

We load the hypercube of each of the media of the database. Then we keep only the countries or territories that fit with the map.  

```{r}
# Check media list
listmedia <- list.files("sirius")
k<-length(listmedia)

# Load first media
media<-listmedia[1]
hc<-readRDS(paste0("sirius/",media,"/hc_int.RDS"))

# Load other media
for (i in 2:k) {
  media<-listmedia[i]
  hc2<-readRDS(paste0("sirius/",media,"/hc_int.RDS"))
  hc<-rbind(hc,hc2)
}
rm(hc2)
class(hc)


# Filter carto
map<-readRDS("geom/gravitymap_V5.rds") %>% filter(is.na(last_year)==T)


# Extract name and code
codename<-map %>% st_drop_geometry() %>% select(ISO3=iso3, country)
mystates<-names(table(map$iso3))

hc <- hc %>% filter(where1 %in% mystates, 
                    where2 %in% mystates,
                    where1 !="PRI",
                    where2 != "PRI")

```

### Matrix of salience by host countries

Finally we standardize in order to obtain a salience of frequency per 10,000 news. 

```{r}
# Agregate
tab<-hc[,.(  Fij=sum(news)),
        .(i=who,j=where1)]

# Transform in matrix
tab2<-dcast(tab, formula = (j~i),fill = 0)
#write.table(tab2,"coverage.csv",row.names = F, sep=";")
mat<-as.matrix(tab2[,-1])
row.names(mat)<-tab2$j

# Put host-guest cells to 0

n<-dim(mat)[1]
p<-dim(mat)[2]
rowN<-row.names(mat)
colN<-substr(colnames(mat),1,3)
for (i in 1:n) {
  for (j in 1:p) {
    
    if (rowN[i]==colN[j]) {mat[i,j]<-0}
  }
}


## Compute value p. 100000

t<-apply(mat,2,sum,na.rm=T)
mat <- round(100000*prop.table(mat,2))


# Put host-guest cells to 0

n<-dim(mat)[1]
p<-dim(mat)[2]
rowN<-row.names(mat)
colN<-substr(colnames(mat),1,3)
for (i in 1:n) {
  for (j in 1:p) {
    
    if (rowN[i]==colN[j]) {mat[i,j]<-NA}
  }
}

# Transform in dataframe
tab<-as.data.frame(mat)
tab$ISO3 <- substr(rownames(mat),1,3)


# Add name


tab <- tab %>% left_join(codename)
tab<-tab[,c(25,26,1:24)]


kable(tab,caption = "Salience of countries by media (p. 100,000)")
```

### Compute parameters

```{r}
## Add parameters
tab$mean <- apply(mat,1,mean, na.rm=T)
tab$rnk<-rank(-tab$mean)
tab$sd<-apply(mat,1,sd,na.rm=T)
tab$min <- apply(mat, 1,min, na.rm=T)
tab$Q1 <- apply(mat, 1, quantile, 0.25, na.rm=T)
tab$Q2 <- apply(mat,1,median, na.rm=T)
tab$Q3 <- apply(mat, 1, quantile, 0.75, na.rm=T)
tab$max<-apply(mat,1,max,na.rm=T)
tab$nbsup0 <- apply(mat>0,1,sum,na.rm=T)
tab$nbsup10 <- apply(mat>10,1,sum,na.rm=T)
tab$nbsup100 <- apply(mat>100,1,sum,na.rm=T)

param<-tab[,c(28, 1,2,27,29:37)] %>% arrange(-mean)


kable(param, digits = 0,row.names = F, caption = "Salience of countries (p. 100,000)")

```



## Typology of media profiles


### Correlation matrix

We establish a correlation matrix based on the Spearman correlation coefficient based on rank. The rank correlation has been chosen against the classical Pearson coefficient because it is less influenced by the weight of top countries (USA, China, Russia, ...) and is more likely to reval regional organisation of news at world scale. 

The correlation are based after exclusion of the host countries (where media are located) . For example, the correlation between *Le Monde* and *Times of India* will exclude the salience of France and India from the vectors of values used for the measure of correlation. 

```{r}

mat2 <- tab[tab$nbsup10>5,3:26]

# Correlation matrix (pearson)
#cor1 <- cor(mat2,use = "pairwise.complete.obs")


# Correlation matrux (Spearman)
cor2 <- cor(mat2,use = "pairwise.complete.obs", method="spearman")


kable(cor2, digits=2, caption = "Spearman correlation coefficient between median values")
```



### Principal Component Analysis

The PCA provide a first viusalization of the proximity between media and some clustrs can be easily observed on the plan defined by the two first component. It is a simple approximation and better results could be obtained with Multi Dimensional Scaling (MDS) or Network Analysis (it will be done later).

```{r}

#library(explor)
acp<-PCA(cor2, graph=F)
#explor(acp)
res <- explor::prepare_results(acp)
explor::PCA_ind_plot(res, xax = 1, yax = 2, ind_sup = FALSE, lab_var = "Lab",
    ind_lab_min_contrib = 0, col_var = NULL, labels_size = 9, point_opacity = 0.5,
    opacity_var = NULL, point_size = 64, ellipses = FALSE, transitions = TRUE,
    labels_positions = "auto", xlim = c(-8.36, 5.68), ylim = c(-7.93, 6.11))

```

- **Comment** : The figure reveals immediately a clustering of media based on geographical or language proximity. Notice the specific situation of the french media *Le Monde* in an intermediate position between media from Western Europe and media from Northern Africa. 




### Hierarchical Cluster Analysis

The hierarchical cluster analysis is realized on the matrix of dissimilarity between $D$ defined as the difference between 1 and the Spearman correlation coefficient : 

$D_{ij} = 1 - \rho_{ij}$

We can then apply different clustering methods. We have chosen here the classical method of Ward. 

```{r}
mydis<-1-cor2

cah_euc <- HCPC(acp,nb.clust = 7,method = "ward",graph = F)

plot(cah_euc,choice = "tree",title = "Hierarchical clustering of media",)


```

- **Comment : ** The tree reveals a first partition in three main groups A, B and C, followed by a subdivision in 7 subgroups. The group A characterized media from Subsaharan Africa (A.1)  and North Africa & Arabia  (A.2). The group B is associated only to coutries located in Latin America .
Finally, the group C is divided in four subgroups corresponding  to Black Sea region (B.1),  Asia-Pacifica (B.2), Northern America and the U.K (B.3) and Western Europe (B.4).


